---
title: "Homework 6"
author: "Vinny Sriram, Lee-Ann Kao, Jennifer Centa"
date: "2024-04-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**We are working with the Wiconsin.Forest.csv loaner dataset, as our own dataset is not appropriate for all the methods of ordination we discussed. The group members are Vinny Sriram, Lee-Ann Kao, and Jennifer Centa. Our emails are vinny.sriram@yale.edu, lee-ann.kao@yale.edu, and jennifer.centa@yale.edu**

```{r}
library(vegan)
library(vegan3d)
library(mgcv)
library(MASS)
library(rgl)
```

### Set-Up
```{r}
#read data
forest <- read.csv("http://reuningscherer.net/multivariate/data/Wisconsin.Forest.csv")

rownames(forest) <-forest[,1]
forestenv <- matrix(forest[,17], ncol = 1)
rownames(forestenv) <- forest[, 1]
colnames(forestenv) <- c("Stage")
forest <- forest[, -c(1, 17)]
forestenv <- data.frame(forestenv)
```


### Question 1: Correspondence Analysis

*We are performing correspondence analysis using the 'cca()' function, looking at the relationship between plots (ordered from early to late stage) and the relative abundance of 14 species of plants in Wisconsin. We have 9 axes (the minimum of 10 plots and 14 species is 10, minus 1 is 9).*
```{r}
#Perform correspondence analysis
forest_cca <- cca(forest)
forest_cca
```


### Question 2: Inertia

*Our total Inertia is 0.7215, meaning that our first CA direction accounts for 0.4506/0.7215 = 62.45% of the total inertia, and our first two CA direction account for around (0.4506 + 0.0899)/0.7215 = 74.91% of the total inertia, or variance in our analysis. This is a decent percentage of variance/inertia explained by our first two CA directions.*
```{r}
#plot results
plot(forest_cca, type = "n" , xlim = c(-2, 2.5))
text(forest_cca, dis = "wa", labels = row.names(forest))
points(forest_cca, pch = 21, col = "red", bg = "yellow", cex = 0.8)
text(forest_cca, "species", col = "blue", cex = 0.6)

#detrended correspondence analysis
forest_dca <- decorana(forest)
plot(forest_dca, type = "n")
text(forest_dca, display = c("sites"), labels = row.names(forest), cex = 0.86)
points(forest_dca, pch = 21, col = "red", bg = "yellow", cex = 0.6)
text(forest_dca, "species", col = "blue", cex = 0.6)
```


### Questions 3 and 4: Data Snaking and Correspondence Analysis Conclusions
*There is evidence of data snaking along the stages, with early stages along the bottom right, traveling up, back down and left across the center, and then ending along the center left at later stages. It makes sense for different species to be more or less prevalent along different ecological stages, rising and falling in frequency rather than all having simple linear relationships with time/stages. From our CCA plots we can see that Shagbark Hickory, Bur Oak, and Black Oak tend to be more prevalent in the pioneer/early stages. Black Cheery, American Elm, White Oak, and Butternut then become more prevalent along the early-middle stages, followed by Red Oak, Black Walnut, and Yellowbud Hickory. Ironwood, Basswood, Sugar Maple, Slippery Elm, and Common name are more prevalent around climax/late stages. Because of this, our data would likely curl in higher-dimensional space as explained in lecture, and so once it is projected down to lower dimensional space there is a "snake-like" winding structure.*

*From our analysis we see a clear pattern following from early to late stages, and identified species that tend to be more prevalent/occur more frequently at difference stages (groupings of which are stated above). We reduced the dimensionality of the data to visualize in two-dimensions/axes which explain around 74.91% of the total inertia/variance, and we found evidence of data snaking in higher dimensional space.*

### Question 5: Multidimensional Scaling
```{r}
forest <- forest [, c(-1:-2, -17)]

forest.mds1 <- metaMDS(forest, k = 1)
forest.mds2 <- metaMDS(forest, k = 2)
forest.mds3 <- metaMDS(forest, k = 3)
forest.mds4 <- metaMDS(forest, k = 4)
forest.mds5 <- metaMDS(forest, k = 5)

plot(forest.mds1, type = "t", cex = 0.6)
plot(forest.mds2, type = "t", cex = 0.6)
plot(forest.mds3, type = "t", cex = 0.6)
plot(forest.mds4, type = "t", cex = 0.6)
plot(forest.mds5, type = "t", cex = 0.6)
```

*Given the forest data describes the relative abundance of each tree species at each sight, Bray-Curtis Dissimilarity is the best  measure for multidimensional scaling.*

### Question 6: Stress and Screeplot

*The stress values decrease with increasing dimensions, as expected, because higher dimensions can better accommodate the data's complexity. For 1-dimensional MDS, we achieved a stress value around 0.07798077, which indicates a reasonable fit, considering that MDS is trying to represent the data in just one dimension. As it was increased to 2 dimensions, the stress drops significantly to around 0.03280863, suggesting a much better fit and likely providing a more meaningful representation of your data. Although the stress continues to drop as number of dimensions is increased, increasing number of dimensions adds to the complexity. Therefore, to limit complexity while preserving a good fit it might make sense to stick with two dimensions, but a screeplot could help further in making this decision. The plots demonstrate a clear different between 1 and 2 dimensions. 2 and 3 dimensions are relatively similar. 4 and 5 dimensions are similar to each other, but have a very different distribution shape to 2 and 3 dimensions*

```{r}
stress <- c(forest.mds1$stress, forest.mds2$stress, forest.mds3$stress, forest.mds4$stress, forest.mds5$stress)
plot(c(1:5), stress, col = "blue", lwd = 3, type = "b", xlab = "Dimensions")
```

*Based on the scree plot, it seems that the elbow is at 3, so it would make sense to stick with the three-dimensional solution.*

### Question 7: Two-Dimensional Plot of MDS Results
```{r}
plot(forest.mds2, main = "2D MDS with Bray-Curtis Dissimilarity", type = "text")
```

### Question 8: Overlay Continuous Variables to Interpret Ordination Axis & Calculate p-values

```{r}
#ordination plot
fig <- ordiplot(forest.mds2, type = "none", cex = 1.1)
text(fig, "species", col = "red", cex = 0.8)
text(fig, "sites", col = "blue", cex = 0.8)

fit <- envfit(forest.mds2, forestenv, permutations = 1000)
plot(fit, col = "black", lwd = 3)
fit   #stage is significant

#wireplot with interpolated surface
fig <- ordiplot(forest.mds2, type = "none", cex = 1.1, main = "NMDS for Forest Data")
text(fig, "species", col = "red", cex = 0.7)
text(fig, "sites", col = "black", cex = 0.7)
plot(fit)
tmp1 <- with(forestenv, ordisurf(forest.mds2, Stage, add = TRUE))

#plot of functions for Stage of Forest
vis.gam(tmp1, main = "Stage")
```

**From the p-value, we can conclude that the stage of the forest is significant. Not just that, but the fact that the vector is long indicates a strong correlation, meaning that changes in the environmental variable (stage) have a greater influence on the distribution of species in the ordination space. This signifies a positive and significant correlation between the stage of the forest and the distribution of species in the NMDS ordination space, indicating that the progression of forest succession is associated with changes in species composition. The wireplot's interpolated surface onto the ordination plot shows how the distribution of forest stages changes, and it identifies regions where specific stages of the forest are more prevalent and how they relate to species composition. For example, from the plots we can see that higher stage forests contain trees such as sugar maple and basswood as opposed to black oak.**

### Question 9: Perform Canonical Correspondence Analysis

```{r}
forestcca <- cca(forest, forestenv, scale = "FALSE")
plot(forestcca)

#get summary information
summary(forestcca)
```


### Question 10: Compare methods used and discuss conclusions reached

**Correspondance Analysis revealed data snaking along succession stages, which shows non-linear relationships between species composition and forest succession stages, suggesting that different species exhibit varying prevalence across ecological stages. MDS in 2D represented the data in lower-dimensional space while preserving the variance, but based on the scree plot suggesting a three-dimensional solution, for further analysis we should consider a three-dimensional MDS. By overlaying the continuous variable of forest stage, the plots clearly show the strong relationship between forest stages and species composition. For this data set, environmental fitting and wireplot provided the clearest visualization of how the distribution of forest stages changes across the ordination space and their association with species composition. **
