---
title: "Homework 6"
author: "Vinny Sriram, Lee-Ann Kao, Jennifer Centa"
date: "2024-04-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**We are working with the Wiconsin.Forest.csv loaner dataset, as our own dataset is not appropraite for all the methods of ordination we discussed The group members are Vinny Sriram, Lee-Ann Kao, and Jennifer Centa. Our emails are vinny.sriram@yale.edu, lee-ann.kao@yale.edu, and jennifer.centa@yale.edu. **

```{r}
library(vegan)
library(vegan3d)
library(mgcv)
library(MASS)
library(rgl)
```

### Set-Up
```{r}
#read data
forest <- read.csv("http://reuningscherer.net/multivariate/data/Wisconsin.Forest.csv")

rownames(forest) <-forest[,1]
forestenv <- matrix(forest[,17], ncol = 1)
rownames(forestenv) <- forest[, 1]
colnames(forestenv) <- c("Stage")
forest <- forest[, -c(1, 17)]
forestenv <- data.frame(forestenv)
```


### Question 1: Correspondence Analysis

*We are performing correspondence analysis using the 'cca()' function, looking at the relationship between plots (ordered from early to late stage) and the relative abundance of 14 species of plants in Wisconsin. We have 9 axes (the minimum of 10 plots and 14 species is 10, minus 1 is 9).*
```{r}
#Perform correspondence analysis
forest_cca <- cca(forest)
forest_cca
```


### Question 2: Inertia

*Our total Inertia is 0.7215, meaning that our first CA direction accounts for 0.4506/0.7215 = 62.45% of the total inertia, and our first two CA direction account for around (0.4506 + 0.0899)/0.7215 = 74.91% of the total inertia, or variance in our analysis. This is a decent percentage of variance/inertia explained by our first two CA directions.*
```{r}
#plot results
plot(forest_cca, type = "n" , xlim = c(-2, 2.5))
text(forest_cca, dis = "wa", labels = row.names(forest))
points(forest_cca, pch = 21, col = "red", bg = "yellow", cex = 0.8)
text(forest_cca, "species", col = "blue", cex = 0.6)

#detrended correspondence analysis
forest_dca <- decorana(forest)
plot(forest_dca, type = "n")
text(forest_dca, display = c("sites"), labels = row.names(forest), cex = 0.86)
points(forest_dca, pch = 21, col = "red", bg = "yellow", cex = 0.6)
text(forest_dca, "species", col = "blue", cex = 0.6)
```


### Questions 3 and 4: Data Snaking and Correspondence Analysis Conclusions
*There is evidence of data snaking along the stages, with early stages along the bottom right, traveling up, back down and left across the center, and then ending along the center left at later stages. It makes sense for different species to be more or less prevalent along different ecological stages, rising and falling in frequency rather than all having simple linear relationships with time/stages. From our CCA plots we can see that Shagbark Hickory, Bur Oak, and Black Oak tend to be more prevalent in the pioneer/early stages. Black Cheery, American Elm, White Oak, and Butternut then become more prevalent along the early-middle stages, followed by Red Oak, Black Walnut, and Yellowbud Hickory. Ironwood, Basswood, Sugar Maple, Slippery Elm, and Common name are more prevalent around climax/late stages. Because of this, our data would likely curl in higher-dimensional space as explained in lecture, and so once it is projected down to lower dimensional space there is a "snake-like" winding structure.*

*From our analysis we see a clear pattern following from early to late stages, and identified species that tend to be more prevalent/occur more frequently at difference stages (groupings of which are stated above). We reduced the dimensionality of the data to visualize in two-dimensions/axes which explain around 74.91% of the total inertia/variance, and we found evidence of data snaking in higher dimensional space.*
